Measurement and Probability in Fuzzy Quantum Theories

### 2016-12-02 prop.tex
prop.tex Copy from proposal\prop.tex

We propose to revisit quantum mechanics, quantum information, and
quantum computation from this resource-aware perspective. Our initial
results in that domain showed how subtle the issues can be~\cite{usat}:
a straightforward replacement of the complex numbers by a finite field
yields a variant of quantum mechanics in which computationally hard
problems like UNIQUE-SAT (which decides whether a given Boolean
formula is unsatisfiable or has exactly one satisfying assignment) can
be deterministically solved in constant time. To eliminate such
unrealistic theories requires delicate analysis of the structure of
the Hilbert space, the process of observation, and the notion of
probability teasing apart their reliance on the infinitely precise
real numbers~\cite{geometry2013,DQT2014}. In this proposal, our aim is
to shift focus from the infinitely-specified but not directly
observable quantum states, to observable measurable properties of
quantum systems and their probabilities. Furthermore, we insist that
our theories of measurement and probability only refer to finitely
communicable evidence within feasible computational bounds. It follows
that states, observations, and probabilities all become ``fuzzy'',
i.e., specified by intervals of confidence that can only increase in
precision if the available resources increase proportionally. Our
notion of ``fuzzy quantum mechanics'' is related to existing
work~\cite{GranikCaulfield1996,Pykacz2013,SNL2009,Gudder2005,aerts1993physical}
but, as will be explained in more detail, is distinguished by its
computational character.

@@

In particular, the objectives to be achieved include:
\begin{itemize}
\item A computational framework for quantum measurement where quantum
  systems are described in terms of interactive processes. The
  framework takes care of the resources required (including the ones
  related to the observer) to determine a given property of the
  measured system.
\item Develop a measurement framework based on quantum interval-valued
  probabilities measures.
\item Assess the validity of fundamental theorems of quantum
  mechanics, such as Gleason, Bell, and Kochen-Specker, in a quantum
  formulation stripped away of infinite resources. We will aim to
  provide a definite answer to the Meyer-Mermin
  debate~\cite{PhysRevLett.83.3751, Mermin1999, BarrettKent2004} on
  the impact of finite precision measurements on the relevance of the
  Kochen-Specker
  theorem~\cite{kochenspecker1967,Redhead1987-REDINA,peres1995quantum,Jaeger2007}.
\item Study the computational complexity of quantum theories and
  simulations based on discrete (computable) fields.
\end{itemize}
We argue that no matter what the results are their impact will be
strong. Positive results that formulate a computable, with clear
resource bounds, theory of quantum measurement and quantum probability
would be essential for understanding and realizing quantum
computation. Any negative results would redirect research that aims at
realizing quantum computation to other approaches. At an intellectual
level, the results have the potential to solve or clarify several
paradoxes in quantum mechanics, quantum information, computability,
and complexity theory.

@@

\noindent It is useful to think that this probability measure is
completely determined by the ``reality'' of the two coins in question
and their characteristics, in the sense that each pair of coins
induces a measure, and each measure must correspond to some pair of
coins. The measure above would be induced by two particular coins such
that the first coin is twice as likely to land tails up than heads up
and the second coin is double-headed. In a strict computational
or experimental setting, one should question the reliance of the
definition of probability space on the uncountable and
uncomputable real
interval~$[0,1]$~\cite{Turing_1937,Ziegler2007,weihrauch2012computable}.
This interval includes numbers like
$0.h_{1}h_{2}h_{3}\ldots$ where $h_{i}$ is 1 or 0 depending on whether
Turing machine $M_{i}$ halts or not. Such numbers cannot be
computed. This interval also includes numbers like $\frac{\pi}{4}$
which can only be computed with increasingly large resources as the
precision increases. Therefore, in a resource-aware computational or
experimental setting, it is more appropriate to consider probability
measures that map events to a set of elements computable with a fixed
set of resources. We expand on this observation in the next section
and then recall its formalization using interval-valued probability
measures~\cite{Weichselberger2000,JamisonLodwick2004}.

@@

In the previous example, we assumed the probability~$\pmeas(E)$ of
each event~$E$ is known a priori. In reality, although each event is
assumed to have a probability, the exact value of $\pmeas(E)$ may not
be known. According to the \emph{frequency interpretation of
  probability} (which we will revisit when moving to the quantum
case)~\cite{Venn1876,Hajek2012}, 
to determine the probability of an event, we run $N$
independent trials which gives us an approximation of the (assumed)
``true'' or ``real'' probability. Let $x_{i}$ be 1 or 0 depending on
whether the event~$E$ occurs in the $i$-th trial or not, then
$\pmeas(E)$ could be approximated to given accuracy~$\epsilon>0$ by
the relative frequency~$\frac{1}{N}\sum_{i=1}^{N}x_{i}$ with the
probability converging to one as $N$ goes to infinity, i.e.,
\[
\forall\epsilon>0,\lim_{N\rightarrow\infty}\pmeas\left(\left|\pmeas(E)-\frac{1}{N}\sum_{i=1}^{N} x_{i}\right|<\epsilon\right)=1\textrm{ .}
\]
This fact is called the law of large numbers~\cite{Bernoulli2006,Kolmogorov1950,Uspensky1937,Shafer1976,544199}.

Let's look at a concrete example. Suppose we drop a needle of length
$\ell$ onto a floor made of equally spaced parallel lines a distance
$h$ apart, where $\ell<h$. It is a known fact that the probability of
the needle crossing a line is
$\frac{2\ell}{\pi
  h}$~\cite{Buffon1777,DeMorgan1872,Hall1873,Uspensky1937}.
Consider an experimental setup consisting of a collection of $N$
identical needles of length $\ell$. We throw the $N$ needles one
needle at a time, and observe the number $M$ of needles that cross a
line, thus estimating the probability of a needle crossing a line to
be $\frac{M}{N}$. In an actual experiment with $500$ needles and the
ratio $\frac{\ell}{h}=0.75$~\cite{Hall1873}, it was found that $236$
crossed a line so the relative frequency is $0.472$ whereas the
idealized mathematical probability is $0.4774\ldots$.  In a larger
experiment with $5000$ needles and the ratio
$\frac{\ell}{h}=0.8$~\cite{Uspensky1937}, the relative frequency was
calculated to be $0.5064$ whereas the idealized mathematical
probability is $0.5092\ldots$. We see that the observed probability
approaches $\frac{2\ell}{\pi h}$ but only if \emph{larger and larger
  resources} are expended. These resource considerations suggest that
it is possible to replace the real interval $[0,1]$ with rational
numbers up to a certain precision related to the particular experiment
in question. There is clearly a connection between the number of
needles and the achievable precision: in the hypothetical experiment
with 3 needles, it is not sensible to retain 100 digits in the
expansion of $\frac{2\ell}{\pi h}$.

There is another more subtle assumption of unbounded computational
power in the experiment. We are assuming that we can always determine
with certainty whether a needle is crossing a line. But ``lines'' on
the the floor have thickness, their distance apart is not exactly $h$,
and the needles' lengths are not all absolutely equal to $\ell$.
These perturbations make the events ``fuzzy.'' Thus, in an experiment
with limited resources, it is not possible to talk about the idealized
event that exactly $M$ needles cross lines as this would require the
most expensive needles built to the most precise accuracy, laser
precision for drawing lines on the floor, and the most powerful
microscopes to determine if a needle does cross a line. Instead we
might talk about the event that $M-\delta$ needles evidently cross
lines and $M+\delta'$ needles plausibly cross lines where $\delta$ and
$\delta'$ are resource-dependent approximations. This fuzzy notion of
events leads to probabilities being only calculable within intervals
of confidence reflecting the certainty of events and their
plausibility. This is indeed consistent with published experiments: in
an experiment with $3204$ needles and the ratio
$\frac{\ell}{h}=0.6$~\cite{DeMorgan1872}, $1213$ needles clearly
crossed a line and $11$ needles were close enough to plausibly be
considered as crossing the line: we would express the probability in
this case as the interval
$\left[\frac{1213}{3204},\frac{1224}{3204}\right]$ expressing that we
are certain that the event has probability at least
$\frac{1213}{3204}$ but it is possible that it would have probability
$\frac{1224}{3204}$.

@@

As motivated above, an event $E_{1}$ may have an interval of
probability $[l_{1},r_{1}]$. Assume that another disjoint event
$E_{2}$ has an interval of probability $[l_{2},r_{2}]$, what is the
interval of probability for the event $E_{1}\cup E_{2}$? The answer is
somewhat subtle: although it is possible to use the sum of the
intervals $[l_{1}+l_{2},r_{1}+r_{2}]$ as the combined probability, one
can find a much tighter interval if information \emph{against} the
event (i.e., information about the complement event) is also taken
into consideration. Formally, for a general event $E$ with probability
$[l,r]$, the evidence that contradicts $E$ is evidence supporting the
complement of $E$.  The complement of $E$ must therefore have
probability $\left[1-r,1-l\right]$ which we abbreviate
$\left[1,1\right]-\left[l,r\right]$.  Given a sample space~$\Omega$
and its set of events~$\events$, a
function~$\bar{\mu}:\events\rightarrow[0,1]$ is a classical
interval-valued probability measure if and only if $\bar{\mu}$
satisfies the following conditions~\cite{JamisonLodwick2004} where the
last line uses $\subseteq$ to allow for tighter intervals that exploit
the complement event:
\begin{itemize}
\item $\bar{\mu}(\emptyset)=[0,0]$.
\item $\bar{\mu}(\Omega)=[1,1]$. 
\item For any event $E$,
  $\bar{\mu}\left(\Omega\backslash E\right)=\left[1,1\right]-\bar{\mu}\left(E\right)$
\item For a collection $\left\{ E_{i}\right\} _{i=1}^{N}$ of pairwise
  disjoint events, we have
  $\bar{\mu}\left(\bigcup_{i=1}^{N}E_{i}\right)\subseteq\sum_{i=1}^{N}\bar{\mu}\left(E_{i}\right)$.
\end{itemize}

@@

\item If we insist that probabilities cannot be infinitely precise and
  are bound to be approximations represented by intervals of
  confidence, then quantum states themselves cannot be infinitely precise
  because Gleason's theorem is no longer applied. 
  This raises an important philosophical
  question with immense practical relevance in the context of quantum
  computing. The question, which we could largely bypass in the
  classical case, is whether there is, in the background of all the
  measurements and probabilities, a ``real'' infinitely-precise
  quantum state that is being approximated and that, in principle, can
  be more and more accurately described given more and more
  resources. In other words, does a Gleason,
  Bell~\cite{BellBook1987,Redhead1987-REDINA,peres1995quantum,Jaeger2007}, 
  or Kochen-Specker
  theorem still hold in a world with finite resources?  The answer
  will have to be \emph{no} if we are to respect these major theorems
  of quantum mechanics and quantum information. Thus our first radical
  departure from the conventional case will be that the quantum state
  itself must be fuzzy in some sense to be formalized.

@@

The early formalization of the process of computing using Turing
machines~\cite{Turing_1937,blum2012complexity} or the
$\lambda$-calculus~\cite{Church1932,Barendregt:Lambda} viewed a computation as
realizing a mathematical function. From our perspective, the most
important consequence of this view is that the properties of the
computed function existed independently of any observation,
calculation, or interaction. For example, a function is
\emph{injective} or not independently of any observation. An agent can
only interact with the computation by asking questions regarding its
input-output behavior and although such interaction would reveal
fragments of the computed function and its innate properties it could
not change it or affect it in any way.

@@

It is unlikely that there will be a simple a resolution to this
awkward situation. It is more likely that the resolution will emerge
from deep and careful analyses of the foundations of each field. In
previous work, we examined the Hilbert space mathematical framework of
quantum mechanics and teased apart its various aspects that directly
rely on the infinite-precision real or complex numbers. We note two
remarkable earlier results:
\begin{itemize}
\item UNIQUE-SAT: Schumacher and
  Westmoreland~\cite{Schumacher2012-SCHMQT} argue that much of the
  structure of traditional quantum mechanics is maintained in the
  presence of finite fields. In particular, they establish that the
  quantum theory based on the finite field of booleans retains the
  following characteristics of quantum mechanics: the notions of
  superposition, interference, entanglement, and mixed states of
  quantum systems; the time evolution of quantum systems using
  invertible linear operators; the complementarity of incompatible
  observables; the exclusion of local hidden variable theories and the
  impossibility of cloning quantum states; and the presence of natural
  counterparts of quantum information protocols such as superdense
  coding and teleportation. However we have argued that this theory is
  not physically realizable as it allows the solution of
  computationally difficult problems like
  UNIQUE-SAT~\cite{Valiant198685} in constant time~\cite{usat}.

@@

A long term, ultimate, goal of this research proposal would be to
combine the various pieces into a coherent computational model of
quantum mechanics that keeps explicit track of resources needed to
represent the interactive system representing the quantum state, its
evolution, the observables, and the probabilities. Such a
computational model of quantum mechanics would encompass classical
(reversible) computation and potentially revolutionize quantum
thinking with finite resources. The explicit manipulation of resources
enables one of key idea that emerged in our previous work: the
resources used to represent the quantum system are in principle
unrelated to the resources available to the various observers. We
conjecture that this asymmetry may prove key to resolving some of the
quantum paradoxes.  The nature of quantum mechanics realized in terms
of discrete fields and finitely-communicable probability estimates is
amenable to direct computational exploration for reasonable field
sizes and observational accuracy. We have established a set of utility
libraries in Mathematica that supports a wide variety of such
numerical experiments. Concurrently to this effort, we are also
building on our previous work~\cite
{VDS2009,AGVS2007,AGVS2005,VAS2006,finiteQC,Sabry:2003:MQC:871895.871900}
to build semantically well-founded implementations of the
computational models investigated in this proposal.

\paragraph*{Research Questions.} The aim of these implementations
include:
\begin{itemize}
\item Provide an experimental testbed to simulate physical phenomena
  (e.g.~\cite {OrtizEtAl2001,SommEtAl2002}) that complements the
  Mathematica libraries in which quantum algorithms can be simulated;
\item Establish connections between quantum programming concepts and
  classical programming concepts. We plan to make our software
  available for educational and research purposes; and 
\item Analyze the computational complexity of quantum theories and
  simulations based on discrete fields. What other (apart from
  UNIQUE-SAT) supernatural algorithms can be implemented in
  computational models based on discrete fields?
\end{itemize}

@@

\paragraph*{PI Team.} Over the past several years, the PIs, together
with other IU faculty members and students, have formed a group whose
main interest is the study of paradigms of computing that either arise
in nature or are inspired by natural phenomena. The PIs are especially
interested in how the laws of physics can inspire and guide our
abstract computational models, as well as exposing new insights into
the computational framework behind the natural world and the
computability of physical theory. Our group has been active in several
areas related to the theory behind quantum computation. Particular
areas include: mathematical foundations of quantum computation and
complexity~\cite{AGVS2007,VAS2006}, adiabatic quantum computing,
quantum logic, classical and quantum reversible programming
languages~\cite{Sabry:2003:MQC:871895.871900}, analog computation and
its limits, and optical computing. During the past several years, our
discussions have revolved around formulating variants of quantum
mechanics that are computable and that are as close to possible to
classical computing. Several preliminary results have emerged from our
collaboration~\cite{usat,finiteQC,geometry2013,DQT2014}, and the group
is very excited about the potential implications of those
results. This proposal brings forth a particular aspect of this
programme related to its applications to problem areas that have
traditionally been outside the conceptual framework of quantum
mechanics and programming language research. This proposal, if funded,
will give our group more recognition within the university which, at
this point, is committed to generously funding several ``emerging
areas of research''. With the establishment of a new Intelligent
Systems Engineering at Indiana University with a focus nanoscale
sensors and quantum engineering and the recent hires in Physics in
experimental atomic physics, ion-trap quantum simulation, computation,
and measurement, the PI group is in an excellent position to receive
additional funding from the University to reach a critical mass of
quantum information researchers.

@@

\paragraph*{Broader Impacts.} The idea of reversible computation and
its connection to quantum mechanics is at the heart of one of the
deepest mysteries in the universe. Any progress or theory that
explains physical principles computationally is likely to change our
worldview impacting everything from philosophy and religion to all
conventional sciences. On the development of researchers, the project
resulted in the graduation of one Ph.D. student with another one
expected to graduate in Spring 2017, the current supervision of two
additional Ph.D. students, the graduation of an MS student, and the
hosting of a visiting scholar.

>>>>

We propose to revisit quantum mechanics, quantum information, and
quantum computation from this resource-aware perspective. Our initial
results in that domain showed how subtle the issues can
be~\cite{usat,geometry2013,DQT2014}: a straightforward replacement of
the complex numbers by a finite field yields a variant of quantum
mechanics in which computationally hard problems like UNIQUE-SAT
(which decides whether a given Boolean formula is unsatisfiable or has
exactly one satisfying assignment) can be deterministically solved in
constant time. To eliminate such unrealistic theories requires
delicate analysis of the structure of the Hilbert space, the process
of observation, and the notion of probability teasing apart their
reliance on the infinitely precise real
numbers~\cite{geometry2013,DQT2014}. In this proposal, our aim is to
shift focus from the infinitely-specified but not directly observable
quantum states, to observable measurable properties of quantum systems
and their probabilities. Furthermore, we insist that our theories of
measurement and probability only refer to finitely communicable
evidence within feasible computational bounds. It follows that states,
observations, and probabilities all become ``fuzzy'', i.e., specified
by intervals of confidence that can only increase in precision if the
available resources increase proportionally. Our notion of ``fuzzy
quantum mechanics'' is related to existing
work~\cite{GranikCaulfield1996,Pykacz2013,SNL2009,Gudder2005,aerts1993physical}
but, as will be explained in more detail, is distinguished by its
unique computational character.

@@

In particular, the objectives to be achieved include:
\begin{itemize}
\item A computational framework for quantum measurement where quantum
  systems are described in terms of interactive processes. The
  framework takes care of the resources required (including the ones
  related to the observer) to determine a given property of the
  measured system.
\item Develop a measurement framework based on quantum interval-valued
  probabilities measures.
\item Assess the validity of fundamental theorems of quantum
  mechanics, such as Gleason, Bell, and Kochen-Specker, in a quantum
  formulation stripped away of infinite resources. We will aim to
  provide a definite answer to the Meyer-Mermin
  debate~\cite{PhysRevLett.83.3751, Mermin1999, BarrettKent2004} on
  the impact of finite precision measurements on the relevance of the
  Kochen-Specker
  theorem~\cite{kochenspecker1967,Redhead1987-REDINA,peres1995quantum,Jaeger2007}.
\item Study the computational complexity of quantum theories and
  simulations based on discrete (computable) fields.
\item Study the problem of finding a complete set of mutually unbiased
  bases
  (SIC-POVM)~\cite{1751-8121-45-8-085306,1751-8121-47-33-335302,Caves2002}
  over finite Galois fields.
\end{itemize}
We argue that no matter what the results are their impact will be
strong. Positive results that formulate a computable, with clear
resource bounds, theory of quantum measurement and quantum probability
would be essential for understanding and realizing quantum
computation. Any negative results would redirect research that aims at
realizing quantum computation to other approaches. At an intellectual
level, the results have the potential to solve or clarify several
paradoxes in quantum mechanics, quantum information, computability,
complexity theory, and perhaps quantum gravity.

@@

\noindent It is useful to think that this probability measure is
completely determined by the ``reality'' of the two coins in question
and their characteristics, in the sense that each pair of coins
induces a measure, and each measure must correspond to some pair of
coins. The measure above would be induced by two particular coins such
that the first coin is twice as likely to land tails up than heads up
and the second coin is double-headed. In a strict computational
or experimental setting, one should question the reliance of the
definition of probability space on the uncountable and
uncomputable real
interval~$[0,1]$~\cite{Turing_1937,Ziegler2007,weihrauch2012computable}.
This interval includes numbers like
$0.h_{1}h_{2}h_{3}\ldots$ where $h_{i}$ is 1 or 0 depending on whether
Turing machine $\mathit{TM}_{i}$ halts or not. Such numbers cannot be
computed. This interval also includes numbers like $\frac{\pi}{4}$
which can only be computed with increasingly large resources as the
precision increases. Therefore, in a resource-aware computational or
experimental setting, it is more appropriate to consider probability
measures that map events to a set of elements computable with a fixed
set of resources. We expand on this observation in the next section
and then recall its formalization using interval-valued probability
measures~\cite{Weichselberger2000,JamisonLodwick2004}.

@@

In the previous example, we assumed the probability~$\pmeas(E)$ of
each event~$E$ is known a priori. In reality, although each event is
assumed to have a probability, the exact value of $\pmeas(E)$ may not
be known. According to the \emph{frequency interpretation of
  probability} (which we will revisit when moving to the quantum
case)~\cite{Venn1876,Hajek2012}, 
to determine the probability of an event, we run $M$
independent trials which gives us an approximation of the (assumed)
``true'' or ``real'' probability. Let $x_{i}$ be 1 or 0 depending on
whether the event~$E$ occurs in the $i$-th trial or not, then
$\pmeas(E)$ could be approximated to given accuracy~$\epsilon>0$ by
the relative frequency~$\frac{1}{M}\sum_{i=1}^{M}x_{i}$ with the
probability converging to one as $M$ goes to infinity, i.e.,
\[
\forall\epsilon>0,\lim_{M\rightarrow\infty}\pmeas\left(\left|\pmeas(E)-\frac{1}{M}\sum_{i=1}^{M} x_{i}\right|<\epsilon\right)=1\textrm{ .}
\]
This fact is called the law of large numbers~\cite{Bernoulli2006,Kolmogorov1950,Uspensky1937,Shafer1976,544199}.

Let's look at a concrete example. Suppose we drop a needle of length
$\ell$ onto a floor made of equally spaced parallel lines a distance
$h$ apart, where $\ell<h$. It is a known fact that the probability of
the needle crossing a line is
$\frac{2\ell}{\pi
  h}$~\cite{Buffon1777,DeMorgan1872,Hall1873,Uspensky1937}.
Consider an experimental setup consisting of a collection of $M$
identical needles of length $\ell$. We throw the $M$ needles one
needle at a time, and observe the number $M_c$ of needles that cross a
line, thus estimating the probability of a needle crossing a line to
be $\frac{M_c}{M}$. In an actual experiment with $500$ needles and the
ratio $\frac{\ell}{h}=0.75$~\cite{Hall1873}, it was found that $236$
crossed a line so the relative frequency is $0.472$ whereas the
idealized mathematical probability is $0.4774\ldots$.  In a larger
experiment with $5000$ needles and the ratio
$\frac{\ell}{h}=0.8$~\cite{Uspensky1937}, the relative frequency was
calculated to be $0.5064$ whereas the idealized mathematical
probability is $0.5092\ldots$. We see that the observed probability
approaches $\frac{2\ell}{\pi h}$ but only if \emph{larger and larger
  resources} are expended. These resource considerations suggest that
it is possible to replace the real interval $[0,1]$ with rational
numbers up to a certain precision related to the particular experiment
in question. There is clearly a connection between the number of
needles and the achievable precision: in the hypothetical experiment
with 3 needles, it is not sensible to retain 100 digits in the
expansion of $\frac{2\ell}{\pi h}$.

There is another more subtle assumption of unbounded computational
power in the experiment. We are assuming that we can always determine
with certainty whether a needle is crossing a line. But ``lines'' on
the the floor have thickness, their distance apart is not exactly $h$,
and the needles' lengths are not all absolutely equal to $\ell$.
These perturbations make the events ``fuzzy.'' Thus, in an experiment
with limited resources, it is not possible to talk about the idealized
event that exactly $M_c$ needles cross lines as this would require the
most expensive needles built to the most precise accuracy, laser
precision for drawing lines on the floor, and the most powerful
microscopes to determine if a needle does cross a line. Instead we
might talk about the event that $M_c-\delta$ needles evidently cross
lines and $M_c+\delta'$ needles plausibly cross lines where $\delta$ and
$\delta'$ are resource-dependent approximations. This fuzzy notion of
events leads to probabilities being only calculable within intervals
of confidence reflecting the certainty of events and their
plausibility. This is indeed consistent with published experiments: in
an experiment with $3204$ needles and the ratio
$\frac{\ell}{h}=0.6$~\cite{DeMorgan1872}, $1213$ needles clearly
crossed a line and $11$ needles were close enough to plausibly be
considered as crossing the line: we would express the probability in
this case as the interval
$\left[\frac{1213}{3204},\frac{1224}{3204}\right]$ expressing that we
are certain that the event has probability at least
$\frac{1213}{3204}$ but it is possible that it would have probability
$\frac{1224}{3204}$.

@@

As motivated above, an event $E_{1}$ may have an interval of
probability $[l_{1},r_{1}]$. Assume that another disjoint event
$E_{2}$ has an interval of probability $[l_{2},r_{2}]$, what is the
interval of probability for the event $E_{1}\cup E_{2}$? The answer is
somewhat subtle: although it is possible to use the sum of the
intervals $[l_{1}+l_{2},r_{1}+r_{2}]$ as the combined probability, one
can find a much tighter interval if information \emph{against} the
event (i.e., information about the complement event) is also taken
into consideration. Formally, for a general event $E$ with probability
$[l,r]$, the evidence that contradicts $E$ is evidence supporting the
complement of $E$.  The complement of $E$ must therefore have
probability $\left[1-r,1-l\right]$ which we abbreviate
$\left[1,1\right]-\left[l,r\right]$.  Given a sample space~$\Omega$
and its set of events~$\events$, a
function~$\bar{\mu}:\events\rightarrow[0,1]$ is a classical
interval-valued probability measure if and only if $\bar{\mu}$
satisfies the following conditions~\cite{JamisonLodwick2004} where the
last line uses $\subseteq$ to allow for tighter intervals that exploit
the complement event:
\begin{itemize}
\item $\bar{\mu}(\emptyset)=[0,0]$.
\item $\bar{\mu}(\Omega)=[1,1]$. 
\item For any event $E$,
  $\bar{\mu}\left(\Omega\backslash E\right)=\left[1,1\right]-\bar{\mu}\left(E\right)$
\item For a collection $\left\{ E_{i}\right\} _{i=1}^{M}$ of pairwise
  disjoint events, we have
  $\bar{\mu}\left(\bigcup_{i=1}^{M}E_{i}\right)\subseteq\sum_{i=1}^{M}\bar{\mu}\left(E_{i}\right)$.
\end{itemize}

@@

\item If we insist that probabilities cannot be computed to infinite
  precision and are bound to be approximations represented by
  intervals of confidence, then quantum states themselves can only be
  discussed within intervals of confidence. This raises an important
  philosophical question with immense practical relevance in the
  context of quantum computing. The question, which we could largely
  bypass in the classical case, is whether there is, in the background
  of all the measurements and probabilities, a ``real''
  infinitely-precise quantum state that is being approximated and
  that, in principle, can be more and more accurately described given
  more and more resources. In other words, does a Gleason,
  Bell~\cite{BellBook1987,Redhead1987-REDINA,peres1995quantum,Jaeger2007},
  or Kochen-Specker theorem still hold in a world with finite
  resources?  The answer will have to be \emph{no} if we are to
  respect these major theorems of quantum mechanics and quantum
  information. Thus our first radical departure from the conventional
  case will be that the quantum state itself must be fuzzy in some
  sense to be formalized.

@@

The early formalization of the process of computing using Turing
machines~\cite{Turing_1937,blum2012complexity} or the
$\lambda$-calculus~\cite{Church1932,Barendregt:Lambda} viewed a
computation as realizing a mathematical function. From our
perspective, the most important consequence of this view is that the
properties of the computed function existed independently of any
observation, calculation, or interaction. For example, a function is
\emph{injective} or not independent of any observation. An agent can
only interact with the computation by asking questions regarding its
input-output behavior and although such interaction would reveal
fragments of the computed function and its innate properties, it could
not change it or affect it in any way.

@@

It is unlikely that there will be a simple a resolution to this
awkward situation. It is more likely that the resolution will emerge
from deep and careful analyses of the foundations of each field. In
previous work, we examined the Hilbert space mathematical framework of
quantum mechanics and teased apart its various aspects that directly
rely on the infinite-precision real or complex numbers. We note two
remarkable earlier results and an open question:
\begin{itemize}
\item UNIQUE-SAT: Schumacher and
  Westmoreland~\cite{Schumacher2012-SCHMQT} argue that much of the
  structure of traditional quantum mechanics is maintained in the
  presence of finite fields. In particular, they establish that the
  quantum theory based on the finite field of booleans retains the
  following characteristics of quantum mechanics: the notions of
  superposition, interference, entanglement, and mixed states of
  quantum systems; the time evolution of quantum systems using
  invertible linear operators; the complementarity of incompatible
  observables; the exclusion of local hidden variable theories and the
  impossibility of cloning quantum states; and the presence of natural
  counterparts of quantum information protocols such as superdense
  coding and teleportation. However we have argued that this theory is
  not physically realizable as it allows the solution of
  computationally difficult problems like
  UNIQUE-SAT~\cite{Valiant198685} in constant time~\cite{usat}.

@@

\item An important special case of a generalized measurement is the
  symmetric, informationally complete, positive operator valued
  measure (SIC-POVM), which has applications in quantum
  cryptography. This problem has been formulated in the context of
  standard quantum theory where fields are complex numbers.  Although
  a rigorous proof for the existence of SIC-POVMs for arbitrary
  Hilbert space dimensions is still missing, it would be illuminating
  to look for a formulation of the same problem in a discrete quantum
  theory defined in terms of finite Galois fields.
\end{itemize}

A long term, ultimate, goal of this research proposal would be to
combine the various pieces into a coherent computational model of
quantum mechanics that keeps explicit track of resources needed to
represent the interactive system representing the quantum state, its
evolution, the observables, and the probabilities. Such a
computational model of quantum mechanics would encompass classical
(reversible) computation and potentially revolutionize quantum
thinking with finite resources. The explicit manipulation of resources
enables one of key idea that emerged in our previous work: the
resources used to represent the quantum system are in principle
unrelated to the resources available to the various observers. We
conjecture that this asymmetry may prove key to resolving some of the
quantum paradoxes.  The nature of quantum mechanics realized in terms
of discrete fields and finitely-communicable probability estimates is
amenable to direct computational exploration for reasonable field
sizes and observational accuracy. We have established a set of utility
libraries in Mathematica that supports a wide variety of such
numerical experiments. Concurrently to this effort, we are also
building on our previous work~\cite
{VDS2009,AGVS2005,VAS2006,finiteQC,Sabry:2003:MQC:871895.871900}
to build semantically well-founded implementations of the
computational models investigated in this proposal.

\paragraph*{Research Questions.} The aim of these implementations
include:
\begin{itemize}
\item Provide an experimental testbed to simulate physical phenomena
  (e.g.~\cite {OrtizEtAl2001,SommEtAl2002}) that complements the
  Mathematica libraries in which quantum algorithms can be simulated;
\item Establish connections between quantum programming concepts and
  classical programming concepts. We plan to make our software
  available for educational and research purposes; and 
\item Analyze the computational complexity of quantum theories and
  simulations based on discrete fields. What other (apart from
  UNIQUE-SAT) supernatural algorithms can be implemented in
  computational models based on discrete fields?
\item Determine, through numerical computations, solutions for
  SIC-POVM for small vector space dimensions.
\end{itemize}

@@

\paragraph*{PI Team.} Over the past several years, the PIs, together
with other IU faculty members and students, have formed a group whose
main interest is the study of paradigms of computing that either arise
in nature or are inspired by natural phenomena. The PIs are especially
interested in how the laws of physics can inspire and guide our
abstract computational models, as well as exposing new insights into
the computational framework behind the natural world and the
computability of physical theory. Our group has been active in several
areas related to the theory behind quantum computation. Particular
areas include: mathematical foundations of quantum computation and
complexity~\cite{AGVS2005,VAS2006}, adiabatic quantum computing,
quantum logic, classical and quantum reversible programming
languages~\cite{Sabry:2003:MQC:871895.871900}, analog computation and
its limits, and optical computing. During the past several years, our
discussions have revolved around formulating variants of quantum
mechanics that are computable and that are as close to possible to
classical computing. Several preliminary results have emerged from our
collaboration~\cite{usat,finiteQC,geometry2013,DQT2014}, and the group
is very excited about the potential implications of those
results. This proposal brings forth a particular aspect of this
programme related to its applications to problem areas that have
traditionally been outside the conceptual framework of quantum
mechanics and programming language research. This proposal, if funded,
will give our group more recognition within the university which, at
this point, is committed to generously funding several ``emerging
areas of research''. With the establishment of a new Intelligent
Systems Engineering at Indiana University with a focus nanoscale
sensors and quantum engineering and the recent hires in Physics in
experimental atomic physics, ion-trap quantum simulation, computation,
and measurement, the PI group is in an excellent position to receive
additional funding from the University to reach a critical mass of
quantum information researchers.

@@

\paragraph*{Broader Impacts.} The idea of reversible computation and
its connection to quantum mechanics is at the heart of one of the
deepest mysteries in the universe. Any progress or theory that
explains physical principles computationally is likely to change our
worldview impacting everything from philosophy and religion to all
conventional sciences. The project resulted in the graduation of one
Ph.D. student with another one expected to graduate in Spring 2017,
the current supervision of two additional Ph.D. students, the
graduation of an MS student, and the hosting of a visiting scholar.
